# Numerical Data {#NUMDATA}


## Objectives  

1) Define and use properly in context all new terminology, to include: *scatterplot*, *dot plot*, *mean*, *distribution*, *point estimate*, *weighted mean*, *histogram*, *data density*, *right skewed*, *left skewed*, *symmetric*, *mode*, *unimodal*, *bimodal*, *multimodal*, *variance*, *standard deviation*, *box plot*, *median*, *interquartile range*, *first quartile*, *third quartile*, *whiskers*, *outlier*, *robust estimate*, *transformation*.  

2) In `R`, generate summary statistics for a numerical variable, including breaking down summary statistics by groups.  

3) In `R`, generate appropriate graphical summaries of numerical variables.  

4) Interpret and explain output both graphically and numerically.  


## Homework  

### Problem 1  

**Mammals exploratory**. Data were collected on 39 species of mammals distributed over 13 taxonomic orders.  The data is in the `mammals` data set in the **openintro** package.

a. Using the documentation for the `mammals` data set, report the units for the variable `brain_wt`.   

```{r eval=FALSE}
?mammals
```

```{r eval=FALSE}
help(mammals)
```


b. Using `inspect()` how many variables are numeric?  

```{r warning=FALSE}
inspect(mammals)
```

c. What type of variable is `danger`?

It is a categorical variable indicating how much danger the mammal faces from other animals, on a scale of 1 to 5. 

d. Create a histogram of `total_sleep` and describe the distribution.  

```{r warning=FALSE}
gf_histogram(~total_sleep, data = mammals, binwidth = 2)
```

```{r warning=FALSE}
gf_dens(~total_sleep, data = mammals)
```

The distribution is unimodal and skewed to the right. It appears it is centered around a value of 11.

e. Create a boxplot of `life_span` and describe the distribution.  

```{r warning=FALSE}
gf_boxplot(~life_span, data = mammals)
```

f. Report the mean and median life span of a mammal.  

```{r}
mean(~life_span, data = mammals, na.rm = TRUE)
```

```{r}
median(~life_span, data = mammals, na.rm = TRUE)
```

```{r}
favstats(~life_span, data = mammals, na.rm = TRUE)
```


g. Calculate the summary statistics for `life_span` broken down by `danger`. What is the standard deviation of life span in danger outcome 5? 

```{r}
favstats(life_span ~ danger, data = mammals)
```

The standard deviation of life span in danger outcome 5 is 10.19 years. 


### Problem 2  

**Mammal life spans**. Continue using the `mammals` data set.

a. Create side-by-side boxplots for `life_span` broken down by `exposure`. Note: you will have to change `exposure` to a `factor()`. Report on any findings.  

```{r warning=FALSE}
mammals %>%
  gf_boxplot(life_span ~ factor(exposure))
```

Mammals who are more exposed during sleep have a longer life span. There must be a confounding variable, such as the size of the animal or the `danger` variable.


b. What happened to the median and third quartile in exposure group 4?

```{r}
favstats(life_span ~ factor(exposure), data = mammals)
```

The median and third quartile are equal in exposure group 4. There are a large number of the observed mammals with the same life span in this group.  

c. Using the same variables, create faceted histograms. What are the shortcomings of this plot?

```{r warning=FALSE}
gf_histogram(~life_span|factor(exposure), data = mammals)
```

There is not enough data for each histogram; some of the histograms provide little to no information.

We can try combining this information in a single plot, coloring the plot by `exposure`. 

```{r warning=FALSE}
gf_histogram(~life_span, color = ~factor(exposure), data = mammals)
```

This is awful. We can also try adjusting the fill color. 

```{r warning=FALSE}
gf_histogram(~life_span, fill = ~factor(exposure), data = mammals)
```

This is very difficult to look at and interpret. Let's use density plots.

```{r warning=FALSE}
gf_dens(~life_span|factor(exposure), data = mammals)
```

```{r warning=FALSE}
gf_dens(~life_span, color = ~factor(exposure), data = mammals)
```


Which do you think is the best graph?

d. Create a new variable `exposed` that is a factor with level `Low` if exposure is `1` or `2` and `High` otherwise.

```{r}
mammals <- mammals %>%
  mutate(exposed = factor(ifelse((exposure == 1) | (exposure == 2), 
                                 "Low", "High")))
```


```{r warning=FALSE}
inspect(mammals)
```

e. Repeat part c) with the new `exposed` variable. Explain what you see in the plot. 

```{r warning=FALSE}
gf_histogram(~life_span|exposed, data = mammals)
```

```{r warning=FALSE}
gf_dens(~life_span, color = ~exposed, data = mammals)
```

This plot is much easier to interpret. We see that higher exposure during sleep tends to result in a longer life span. It's a lot easier to view this trend with two groups instead of five. This is a strategy we can use when we have too little data in each group. 


### Problem 3  

**Mammal life spans continued**  

a. Create a scatterplot of life span versus length of gestation.  

```{r warning=FALSE}
mammals %>%
  gf_point(life_span ~ gestation)
```


b. What type of association is apparent between life span and length of gestation?   

It is a weak positive association.

c. What type of association would you expect to see if the axes of the plot were reversed, i.e. if we plotted length of gestation versus life span?  

We would expect to see the same - a weak positive association. Since this is observational data, there is no reason to believe there is a causal relationship just by looking at the data. Switching the axis will preserve the association.  

d. Create the new scatterplot suggested in part c).  

```{r warning=FALSE}
mammals %>%
  gf_point(gestation ~ life_span)
```

e. Are life span and length of gestation independent? Explain your reasoning. 

No, there is an association and it appears to be linear. If the plot looked like a "shotgun" blast, we would consider the variables to be independent. However, remember there may be confounding variables that could impact the association between these variables.  

